Left-linear rules in adhesive categories

### 2024-02-14 icalp/fusione.tex
fine

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of the Double Pushout (DPO)
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  More precisely, we focus on left-linear rules, which
  not only consume, preserve and generate entities, as linear rules
  do, but may ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

\item \emph{Well-Switching rewriting
    systems}\\ Strong Enforcing and existing indepence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

@@

This part is dealt with for well-switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well-switching
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well-switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well-switching
\end{itemize}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered and some new
entities can be generated. Concerning concurrency, the chosen notion
is shift equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Section~4]{Ehrig78}\cite[Chapter 3, Sec. 3.4]{CorradiniMREHL97}.

@@

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switching rewriting systems}, 
where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

On the one hand, they capture most categories of
interest for rewriting. In particular, we identify a class of
left-linear rewriting systems on graph-like structures that are
well-switching and we show that rewriting systems on graphs with
equivalences are so. 
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we show that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Specifically, 
the Church-Rosser theorem is fully recovered.
%for well-switching rewriting systems.

@@

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails,
% : the existence of a consistent permutation is only necessary but
% not sufficient
hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the two properties above hold.
We show that much of the theory can be retained: consistency
of switching still holds, while globality of independence  in a
weaker form only, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.

@@

\begin{notation}
  We stipulate some notational conventions that will be used
  throughout the paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects, so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

@@

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property that $\mathcal{M}$-adhesive categories enjoy is given
by the \emph{Van Kampen condition}
\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We
will recall it and examine some of its consequences. First of all, we
need to introduce some terminology and facts regarding subclasses of
$\mor(\X)$.

@@

We will write $m\colon X\mto Y$ to denote that an arrow $m\colon X\to Y$ belongs to $\mathcal{M}$.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
approach (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ is the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

@@

  \vspace{.1cm}
  \noindent
  \parbox{10cm}{\hspace{15pt}Given two objects $G$ and $H$ and a rule
    $\rho=(l,r)$ in $\R$, a \emph{direct derivation $\mathscr{D}$ from
      $G$ to $H$ applying the rule $\rho$} is a diagram as the one
    aside, in which both squares are pushouts. The arrow $m$ is called
    the \emph{match} of the derivation, while $h$ is its
    \emph{back-match}.  We will denote a direct derivation $\dder{D}$
    between $G$ and $H$ as $\dder{D}\colon G\Mapsto H$. }
  \parbox{3cm}{$\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}&
      H}$}
\end{definition}

@@

Derivations naturally compose: given $\der{D}: G \Mapsto H$ and $\der{E}: H \Mapsto F$, we can consider their composition $\der{D} \cdot \der{E}: G \to F$, defined in the obvious way.

@@

\full{
[Proof in \Cref{propUnique-proof}]
}

@@

In this section we discuss the notion of independence between two
consecutive rewriting steps, a fundamental ingredient of the rewriting
theory, which comes into play when viewing a sequence of steps as a
concurrent computation. We observe that moving from linear to
left-linear rules leads to the failure of various basic properties of
the classical notion of independence used in the linear setting and we
single out a framework in which these can be re-established, possibly
in a weakened form.

@@

Sequential independence is the canonical notion of independence in the DPO approach.
%recalled below.

@@

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$, is depicted in Fig.~\ref{fi:derD}. Note that the numbers in $\rho_1$
  were changed consistently to represent the vertical morphisms.
  Steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent via independence pair $(R_1 \to D_2, L_2 \to D_1)$.
  %that map nodes consistently with their numbering.\todo{P: having names for the graphs would probably help}
  A switch of $\der{D}$ along such independence pair is the derivation
  $\der{E}$ in Fig.~\ref{fi:derE}.

@@

As we mentioned, for linear rewriting systems
%when viewing  derivations as concurrent computations, 
it is canonical to identify
derivations that are equal ``up to switching'', i.e. that differs
only for the order of independent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

% \begin{definition}
%   Consider derivations $\der{D}_i$ for $i \in \interval[0]{n}$
%   related by switches
%   \begin{center}
%     $\der{D}_0 \shift{\nu_1} \der{D}_1 \shift{\nu_2} \ldots \shift{\nu_n}
%     \der{D}_n$
%   \end{center}
%   It will be referred as a \emph{switching sequence}.  
% \end{definition}

Switch equivalence is now defined as the equivalence relating derivations that 
can be obtained one from another by a sequence of switches. Moreover, intermediate graphs can be taken up to isomorphism according to abstraction equivalence.

@@

The fact that sequential independence implies switchability always
holds for linear rules (see \Cref{prop:equi}). The result is so
indispensable that it is not even stated, in the sense that
in the literature switchability is not introduced axiomatically as in
Definition~\ref{def:switch}, but it is based on the explicit
construction of a switch.

@@

  Take the set $S:=\{a,b,c\}$ and endow it with the smallest order
  such that $a\leq b$ and $a\leq c$.  Take also the set $\mathbb{N}$
  ordered by $\geq$, so that $0$ is its maximum.  Let $\X$ be the
  category associated to the order
  $(S, \leq)\oplus (\mathbb{N}, \geq)$, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. Notice that the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a pushout
  because $b$ and $c$ do not have a supremum.  Consider a rewriting
  system whose set of rules $\R$ contains the following %two rules
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ of length $2$ below
  \[
    \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
      \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
      \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
      \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  for $i \in \{0,1\}$ and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback object (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

\begin{restatable}[Local Church-Rosser Theorem]{theorem}{thmChurch}
  \label{thm:church}
  Let $(i_0, i_1)$ be a strong independence pair
  between $\dder{D}_0$ and $\dder{D}_1$. Then $\dder{D}_0$ and
  $\dder{D}_1$ are switchable.
\end{restatable}
\full{
  [Proof in \Cref{thmChurch-proof}]
}

The correspondence between sequential independence and switchability
is fundamental. We name the class of rewriting systems where this property holds.

\begin{definition}[Strong Enforcing rewriting systems]
  A left-linear DPO rewriting system is \emph{strong enforcing} if
  every independence pair between two direct derivations is strong.
\end{definition}

Along the lines of \cite{baldan2011adhesivity} we can identify a large
class of adhesive categories such that all left-linear rewriting
systems over such categories are strong enforcing. The details are in
\Cref{app:fill}. Here it is important only to remark that such class
includes $\cat{Set}$ and it is closed by comma and functor category
constructions.  As such, it includes essentially all categories (e.g.,
presheaves over set) that are typically considered for modelling
purposes. Notably, it contains the category $\cat{Graph}$ of directed
graphs.

Still, there are adhesive rewriting systems which are not strong enforcing, as shown by the following example, which is however somehow pathological.

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a
  non-strong enforcing adhesive rewriting system.%  Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

\begin{remark}
  By \Cref{thm:church} the existence of a strong independence pair
  entails switchability, which in turn entails weak sequential
  independence by construction. Strong Enforcing rewriting systems are exactly
  those rewriting systems in which these three notions coincide.
\end{remark}
%}

As a sanity check, note that a source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

\begin{restatable}{proposition}{propEqui}
  \label{prop:equi}
  Every linear DPO rewriting system $(\X, \R)$ is strong enforcing.
\end{restatable}
\full{
[Proof in \Cref{propEqui-proof}]
}

\subsection{well-switching rewriting systems}
\label{subsec:verytame}

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, i.e.,
the fact that there can be more than one independence pair between
the same derivations. This can hinder the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ would be switch
  equivalent, but this is not acceptable when viewing equivalence
  classes of derivations as concurrent computations: in $\der{D}$ the
  first two steps are not sequential independent, while in $\der{D}'$
  they are, intuitively because in $\der{D}$ rule $\rho_1$ uses the
  node generated by $\rho_0$ (adding a self loop to it), while in
  $\der{D}'$ rule $\rho_1$ uses the node which was in the initial
  graph. From the technical point of view, the property of being
  switch equivalent is not closed by prefix, something which prevents
  to derive a sensible concurrent sematics: if we limit to the first
  two steps, we obatain derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  $\dder{D}_0' \cdot \dder{D}_1'$ are not switch equivalent while
  $\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  $\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ would
  be.
\end{example}

@@

\begin{definition}[Well-switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well-switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence pair between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

>>>>

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting.\todo{[A] direi: ``...
    in the context of the Double Pushout (DPO) Approach to rewriting in adhesive categories.''}
    More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

\item \emph{Well Switching rewriting
    systems}\\ Strong Enforcing and existing indepence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

@@

This part is dealt with for well switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well switching
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well switching
\end{itemize}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently from more recent approaches
like SqPO~\cite{CorradiniHHK06}
or PBPO$^+$~\cite{OverbeekER23}) and to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by what is called the
Church-Rosser theorem of DPO rewriting~\cite[Section~4]{Ehrig78}.

@@

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switchable rewriting systems}, 
where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

On the one hand, they capture most categories of
interest for rewriting. In particular, we single out a class of
left-linear rewriting systems on graph-like structures that are
well-switching and we show that rewriting systems on graphs with
equivalences are so. 
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we show that a concurrency theory of rewriting can
be developed for well-switchable rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Specifically, 
the Church-Rosser theorem is fully recovered.
%for well-switchable rewriting systems.

@@

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, BaldanCHKS06, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switchable systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for our work.

@@

\begin{notation}
  We stipulate some notational conventions that will be used
  throughout the paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects: so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

@@

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property that $\mathcal{M}$-adhesive categories enjoy is given
by the so-called \emph{Van Kampen condition}
\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We
will recall it and examine some of its consequences. First of all, we
need to recall some terminology and facts regarding subclasses of
$\mor(\X)$.

@@

We will use $m\colon X\mto Y$ to denote that an arrow $m\colon X\to Y$ belongs to $\mathcal{M}$.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the ``Double-Pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$, such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ is the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

@@

  \noindent
  \parbox{10cm}{\hspace{15pt}Given two objects $G$ and $H$ and a rule
    $\rho=(l,r)$ in $\R$, a \emph{direct derivation $\mathscr{D}$ from
      $G$ to $H$ applying the rule $\rho$} is a diagram as the one
    aside, in which both squares are pushouts. The arrow $m$ is called
    the \emph{match} of the derivation, while $h$ is its
    \emph{back-match}.  We will denote a direct derivation $\dder{D}$
    between $G$ and $H$ as $\dder{D}\colon G\Mapsto H$. }
  \parbox{3cm}{$\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}&
      H}$}
\end{definition}

@@

Derivations naturally compose: given $\der{D}: G \Mapsto H$ and $\der{E}: H \Mapsto F$ we can consider their composition $\der{D} \cdot \der{E}: G \to F$, defined in the obvious way.

@@

[Proof in \Cref{propUnique-proof}]

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, a fundamental ingredient of the rewriting
theory, which comes into play when viewing a sequence of steps as a
concurrent computation. We observe that moving from linear to
left-linear rules leads to the failure of various basic properties of
the classical notion of independence used in the linear setting and we
single out a framework in which these can be re-established, possibly
in a weakened form.

@@

In the DPO approach the canonical notion of independence is that of sequential independence.
%recalled below.

@@

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$, is depicted in Fig.~\ref{fi:derD}. Note that the numbers in $\rho_1$
  were changed consistently to represent the vertical morphisms.
  Steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent via independence pair $(R_1 \to D_2, L_2 \to D_1)$.
  %that map nodes consistently with their numbering.\todo{P: having names for the graphs would probably help}
  This pair can be used to construct a switch $\der{E}$, shown
  in Fig.~\ref{fi:derE}.

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

\begin{definition}
  \todo{A: ma questa definizione va tolta, no?}
  Consider derivations $\der{D}_i$ for $i \in \interval[0]{n}$
  related by switches
  \begin{center}
    $\der{D}_0 \shift{\nu_1} \der{D}_1 \shift{\nu_2} \ldots \shift{\nu_n}
    \der{D}_n$
  \end{center}
  It will be referred as a \emph{switching sequence}.  
\end{definition}

Switch equivalence is now defined as the equivalence relating derivations which can be obatained one from another by a sequence of switches. Moreover, intermediate graphs can be taken up to isomorphism according to abstraction equivalence.

@@

The fact that sequential indepedence implies switchability always
holds for linear rules (see \Cref{prop:equi}). The result is so
indispensable that it is not even stated, in the sense that
in the literature switchability is not introduced axiomatically as in
Definition~\ref{def:switch}, but it is based on the explicit
construction of a switch.

@@

  Take now the set $S:=\{a,b,c\}$ and endow it with the smallest order
  such that $a\leq b$ and $a\leq c$.  Take also the set $\mathbb{N}$
  ordered by $\geq$, so that $0$ is its maximum.  Let $\X$ be the
  category associated to the order
  $(S, \leq)\oplus (\mathbb{N}, \geq)$, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. Notice that in this category the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a pushout
  because $b$ and $c$ do not have a supremum.  Consider a rewriting
  system whose set of rules $\R$ contains the following two rules
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ of length $2$ below
  \[
    \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
      \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
      \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
      \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback object (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

\begin{restatable}[Local Church-Rosser Theorem]{theorem}{thmChurch}
  \label{thm:church}
  Let $(i_0, i_1)$ be a strong independence pair
  between $\dder{D}_0$ and $\dder{D}_1$. Then $\dder{D}_0$ and
  $\dder{D}_1$ are switchable.
\end{restatable}
[Proof in \Cref{thmChurch-proof}]

The correspondence between sequential independence and switchability
is fundamental.  A very large class of DPO rewriting systems -- along the
lines of \cite{baldan2011adhesivity} -- can be identified where all
independence pairs are strong for left-linear rewriting systems. The details are in 
\Cref{app:fill}. Here it is important only to observe that such class includes $\cat{Set}$ and it is closed by comma and functor category constructions. 
As such, it includes essentially all categories (e.g., presheaves over set) that are typically considered for modelling purposes.

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Strong Enforcing rewriting systems]A left-linear DPO rewriting system is \emph{strong enforcing} if every independence pair between two direct derivations is strong. 
\end{definition}

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a non-strong enforcing
  adhesive rewriting system. Instead every left-linear system in category $\cat{Graph}$ is
  strong enforcing, by the results of Appendix~\ref{app:fill}, summarized in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

\begin{remark}
  By \Cref{thm:church} the existence of a strong independence pair
  entails switchability which, in turn, entails weak sequential
  independence by construction. Strong Enforcing rewriting systems are exactly
  those rewriting systems in which these three notions coincide.
\end{remark}
%}

%\rem{Questo e' quanto gia' asserito sopra, mettere in appendice e riferire}

As a sanity check note that a source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

\begin{restatable}{proposition}{propEqui}
  \label{prop:equi}
  Every linear DPO rewriting system $(\X, \R)$ is strong enforcing.
\end{restatable}
[Proof in \Cref{propEqui-proof}]

\subsection{Well-switchable rewriting systems}\label{subsec:verytame}

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, i.e.,
the fact that there can be more than one independence pairs between
the same derivations. This can hinder the core idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ are switch equivalent, but
  this is not acceptable when viewing equivalence classes of
  derivations as concurrent computations: in $\der{D}$ the first two
  steps are not sequential independent, while in $\der{D}'$ they are,
  intuitively because in $\der{D}$ rule $\rho_1$ uses the node
  generated by $\rho_0$, while in $\der{D}'$ rule $\rho_1$ uses the
  node which was in the initial graph. From the technical point of
  view, the property of being switch equivalent is not closed by
  prefix, something which prevents to derive a sensible concurrent
  sematics: if we limit to the first two steps, derivations
  $\dder{D}_0; \dder{D}_1$ and $\dder{D}_0'; \dder{D}_1'$ are not switch
  equivalent while $\der{D}$ and $\der{D}'$ would be
\end{example}

@@

\begin{definition}[Well switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence pair between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE
  ESEMPI QUI} actually well switching.
%
Roughly speakings, these are possibly hierarchical graphical structures where rules are contrained not to merge elements of top level sorts in the hiearchy (for graphs, nodes can be merged while edges can not).
